---
output: html_document
---
Design and Analysis of Computer Experiments
========================================================


**Jason Hilton and Jakub Bijak** - University of Southampton 

Workshop given for IDEM 112 'Agent-based modelling and simulation'  
Part of the MPIDR [International Advanced Studies in Demography series](http://www.demogr.mpg.de/en/education_career/international_advanced_studies_in_demography_3279/default.htm)

Location: Max Planck Institute for Demographic Research, Rostock  
Date: 30^th^ October 2014

***

#Introduction 

In this workshop, we shall work through examples of a number of the techniques discussed in the preceding lectures. 

All files and supporting information are available on the github page [https://github.com/jasonhilton/Comp_Exp_Workshop](https://github.com/jasonhilton/Comp_Exp_Workshop).


#Part 1: Experimental Designs and Simple Metamodels  



## A First Experiment  
We will start by running some very simple experiments to examine some of the ideas discussed in the lecture. 

We will run netlogo through R using the RNetLogo Library, which you are all familiar with. 
You will need to edit the nl.path variable to point to the folder where netlogo is installed on your machine. 

```{r}
library(RNetLogo)
nl.path<- "C:\\Program Files (x86)\\NetLogo 5.1.0"
NLStart(nl.path, gui=F)
```

Let's load up the model. 

```{r}
model.path <- "/models/Sample Models/Earth Science/Fire.nlogo"

NLLoadModel(paste(nl.path,model.path,sep=""))
```



Recall, we are interested in how our model *inputs* map to *outputs* or *responses*.
We can use 
```{r}
NLCommand("set density 77")
NLCommand("setup")
NLDoCommand(100,"go")
initial<- NLReport("initial-trees")
burned<- NLReport("burned-trees")

```









#Part 2: Uncertainty and Emulation






